### 2013-07-01 Publications/2013-TMI/03-introduction.tex
Fixed some references

\Gls{epi} are known to suffer from geometrical distortions and artifacts
due to, mainly, three sources: the subject motion in between acquiring 
different sampling directions, the induced \emph{Eddy currents} on the scanner 
coils, and, finally, distortions caused by the magnetic susceptibility inhomogeneity
present at tissue interface. \emph{Susceptibility distortions} happen along the 
phase-encoding direction, and are most appreciable in the frontal part of 
the brain due to the air/tissue interface located around the sinuses.
Two implications are associated to this artifact: the signal loss caused on
highly distorted regions, and a significant added difficulty in 
registering with structural images (e.g. T1-weighted).

After correcting for artifacts, the difficulty of performing analyses directly
on the \emph{directionally dependent} raw data is not affordable. Thus, 
\gls{dwi} data is reconstructed using one of a numerous set of readily 
available algorithms. The outcome is a voxel-wise model of diffusion shape 
(\todo[inline]{cite here dti, dsi, hardi, etc.and refs}). On top of the
chosen model, some scalar maps describing tissue features is
computed. For instance, the most popular \gls{dwi}-derived scalar map 
is \gls{fa} which depicts the isotropy of water diffusion inside the brain.
Another widely studied map is the \gls{adc} (or \gls{md} that has a dual definition),
which reflects the intensity of the diffusion effect. Many other
directionally invariant measurements have been defined. For a comprehensive study, 
we refer to \citep{ennis_orthogonal_2006}.

@@

A precise delineation of the \gls{csf}, \gls{gm} and \gls{wm} interfacing surfaces
is required with sub-pixel resolution, because whereas tractography algorithms are
very good at estimating the location of bundles in deep \gls{wm}, they are not
god (yet) at identifying where the tracts project into the \gls{gm} 
\citep{jbabdi_tractography:_2011,craddock_imaging_2013}.
Thus, a \gls{wm} segmentation is required by most of the available tractography 
algorithms to \emph{filter} the resulting tractogram. The \gls{gm}-\gls{wm}
interface is necessary to locate the starting and ending points of the detected
fiber bundles, and \gls{csf}-\gls{wm} surface is critical for pruning spurious 
and discontinued fiber bundles.

A number of methodologies have been proposed for \gls{dwi} segmentation, ranging 
from intensity thresholding to atlas-based segmentation. 
The first approach is performed simply thresholding the \gls{fa} map.
Although this methodology was popular among the premier tractography studies,
they were generally limited to certain regions or significant fiber tracts, and thus,
it cannot be applied in whole-brain tractography. Early approaches to \gls{dwi} segmentation 
include level set formulations using scalar maps of direction invariants derived
from the tensor model \citep{zhukov_level_2003}, directly on the diffusion raw data
\citep{rousson_level_2004}, or finding alternative diffusion representations 
\citep{jonasson_representing_2007}. Even though this latter case was restricted to the extraction of
the corpus callosum from a real dataset, the density of the components of the diffusion tensor
are approximated by multivariate Gaussians for first. Iterative clustering performed on the 
B0 volumes of \gls{dwi} data was proposed by \citep{hadjiprocopis_unbiased_2005}.
Later studies investigated the application of probabilistic frameworks combining mixtures of 
gaussians, \gls{mrf} and labeling fusion techniques \citep{liu_brain_2007} using as features 
widely-used \gls{dwi}-derived scalar maps as \gls{fa} or \gls{md}. A similar framework 
combining co-registered structural information (T1 weighted) with \emph{independent orthogonal 
invariants} derived from the \gls{dwi} tensor model was proposed by \citep{awate_multivariate_2008}. 
Some proposals suggest the use of the raw diffusion data (directionally dependent) to avoid fitting 
a certain model \citep{lu_segmentation_2008}. In \citep{han_experimental_2009}, graph-cuts 
voxel-based techniques are proposed using the most common diffusion tensor derived features.
Further developments of the probabilistic approach have been proposed adding more scalar maps
as features and a more detailed treatment of \gls{pve} \citep{kumazawa_partial_2010}.

@@

Some other retrospective methodologies do not make use of the fieldmaps, 
as explicitly modeling the distortion \citep{andersson_modeling_2001},
registering with (anatomically correct) T2-weighted \gls{mri} 
\citep{kybic_unwarping_2000,studholme_accurate_2000,li_accounting_2007,
tao_variational_2009} (see \autoref{sec:registration}), or acquiring an 
extra B0 image with reversed phase encoding direction 
\citep{andersson_how_2003,holland_efficient_2010}.

A comparison of the diverse correction techniques is found in \citep{wu_comparison_2008}.
This study claims that fieldmap correction methodologies are not entirely accurate
and reliable, even though the method is correct in principle. This conclusion
was later confirmed by \citep{tao_variational_2009}. Additional concerns regarding
fieldmap correction are the requirement of an extra acquisition (that is not always
met or it is impractical), or the accuracy of the measured fieldmap that is sensitive
to various effects (such us respiration, blood flow, etc.). All these factors 
have turned susceptibility distortion in \gls{epi} sequences an active field of 
research for the last 10 years.

>>>>

\Gls{epi} are known to suffer from geometrical distortions and artifacts
due to, mainly, three sources: the subject motion in between acquiring 
different sampling directions, the induced \emph{eddy currents} on the scanner 
coils due to the gradient switching \citep{obrien_3d_2013}, and, finally, 
distortions caused by the magnetic susceptibility inhomogeneity
present at tissue interface. \emph{Susceptibility distortions} happen along the 
phase-encoding direction, and are most appreciable in the frontal part of 
the brain due to the air/tissue interface located around the sinuses.
Two implications are associated to this artifact: the signal loss caused on
highly distorted regions, and a significant added difficulty in 
registering with structural images (e.g. T1-weighted).

After correcting for artifacts, the difficulty of performing analyses directly
on the \emph{directionally dependent} raw data is not affordable. Thus, 
\gls{dwi} data is reconstructed using one of a numerous set of readily 
available algorithms. The outcome is a voxel-wise model of diffusion shape 
(in \citep{daducci_isbi2013_2013} \todo{FIX ref} there is available a comprehensive review of 
state-of-art methodologies). On top of the
chosen model, some scalar maps describing tissue features is generally
computed. For instance, the most popular \gls{dwi}-derived scalar map 
is \gls{fa} which depicts the isotropy of water diffusion inside the brain.
Another widely studied map is the \gls{adc} (or \gls{md} that has a dual definition),
which reflects the intensity of the diffusion effect. Many other
directionally invariant measurements have been defined. For a comprehensive study, 
we refer to \citep{ennis_orthogonal_2006}.

@@

A precise delineation of the \gls{csf}, \gls{gm} and \gls{wm} interfacing surfaces
is required with sub-pixel resolution, because whereas tractography algorithms are
very good at estimating the location of bundles in deep \gls{wm}, they are not
god (yet) at identifying where the tracts project into the \gls{gm} 
(limitations on the termination criteria, \citep{jbabdi_tractography:_2011,
craddock_imaging_2013}).
Thus, a \gls{wm} segmentation is required by most of the available tractography 
algorithms to \emph{filter} the resulting tractogram. The \gls{gm}-\gls{wm}
interface is necessary to locate the starting and ending points of the detected
fiber bundles, and \gls{csf}-\gls{wm} surface is critical for pruning spurious 
and discontinued fiber bundles.

A number of methodologies have been proposed for \gls{dwi} segmentation, ranging 
from intensity thresholding to atlas-based segmentation. 
The first approach is performed simply thresholding the \gls{fa} map.
Although this methodology was popular among the pioneer tractography studies,
they were generally limited to certain regions or significant fiber tracts, and thus,
it cannot be applied in whole-brain tractography. Early approaches to \gls{dwi} segmentation 
include level set formulations using scalar maps of direction invariants derived
from the tensor model \citep{zhukov_level_2003}, directly on the diffusion raw data
\citep{rousson_level_2004}, or finding alternative diffusion representations 
\citep{jonasson_representing_2007}. Even though this latter case was restricted to the extraction of
the corpus callosum from a real dataset, the density of the components of the diffusion tensor
are approximated by multivariate Gaussians for first. Iterative clustering performed on the 
B0 volumes of \gls{dwi} data was proposed by \citep{hadjiprocopis_unbiased_2005}.
Later studies investigated the application of probabilistic frameworks combining mixtures of 
gaussians, \gls{mrf} and labeling fusion techniques \citep{liu_brain_2007} using as features 
widely-used \gls{dwi}-derived scalar maps as \gls{fa} or \gls{md}. A similar framework 
combining co-registered structural information (T1 weighted) with \emph{independent orthogonal 
invariants} derived from the \gls{dwi} tensor model was proposed by \citep{awate_multivariate_2008}. 
Some proposals suggest the use of the raw diffusion data (directionally dependent) to avoid fitting 
a certain model \citep{lu_segmentation_2008}. In \citep{han_experimental_2009}, graph-cuts 
voxel-based techniques are proposed using the most common diffusion tensor derived features.
Further developments of the probabilistic approach have been proposed adding more scalar maps
as features and a more detailed treatment of \gls{pve} \citep{kumazawa_partial_2010}.

@@

Some other retrospective methodologies do not make use of the fieldmaps, 
as explicitly modeling the distortion \citep{andersson_modeling_2001},
registering with (anatomically correct) T2-weighted \gls{mri} 
\citep{kybic_unwarping_2000,studholme_accurate_2000,li_accounting_2007,
tao_variational_2009} (we shall cover these methods in 
\autoref{sec:registration}), or acquiring an 
extra B0 image with reversed phase encoding direction 
\citep{andersson_how_2003,holland_efficient_2010}.

A comparison of the diverse correction techniques is found in \citep{wu_comparison_2008}.
This study claims that fieldmap correction methodologies are not entirely accurate
and reliable, even though the method is correct in principle. The conclusion
was later confirmed by \citep{tao_variational_2009}. Additional concerns regarding
fieldmap correction are the requirement of an extra acquisition (that is not always
met or it is impractical), or the accuracy of the measured fieldmap that is sensitive
to various effects (such us respiration, blood flow, etc.). All these factors 
have turned susceptibility distortion in \gls{epi} sequences an active field of 
research for the last 10 years.

### 2013-07-01 Publications/2013-TMI/04-methods.tex
Fixed some references

\Gls{acwe} and level-set based formulations have been widely applied in image
processing, as reviewed in \citep{suri_shape_2002}. We suggest clustering the 
current methodologies of template-based segmentation methods into three groups. 
The first group typically adds a shape prior term to the energy functional of 
an evolving active contour \citep{bresson_variational_2006,
chan_level_2005,chen_using_2002,cremers_kernel_2006,gastaud_combining_2004}.
These methods generally have a explicit description of the expected relative boundary 
locations of the object to be delineated, and some even model the statistical deviations
from this average shape. By including a coordinates mapping, it is possible to perform
active contours based registration between timesteps in a time-series or between different
images \citep{wyatt_map_2003,paragios_level_2003,vemuri_joint_2003,yezzi_variational_2003}.
This second group is closely related to atlas-based segmentation methods 
\citep{gorthi_segmentation_2009,gorthi_active_2011,pohl_unifying_2005,
pohl_bayesian_2006,wang_joint_2006}, where the prior 
imposes consistent voxel-based classification of contiguous regions.
A comprehensive summary of the \gls{acwe}-derived methodologies with special attention
to joint registration-segmentation methods is found in \citep{gorthi_active_2011}.

Our joint registration and segmentation method is mainly inspired in the active
deformation fields presented by \citep{gorthi_active_2011}, that is a generalization
of the methodologies presented by \citep{bertalmio_morphing_2000,vemuri_joint_2003,
yezzi_variational_2003}.
All these formulations are deduced from the general level set evolution equation
introduced by \citep{osher_fronts_1988}:
\begin{equation}
\frac{\partial \Phi_D(x,t)}{\partial t} = \nu ( \Phi_D(x,t)) \, \left| \nabla \Phi_D(x,t) \right|
\end{equation}
where $\nu$ is the velocity of the flow or speed function that contains the local
segmentation and contour regularization constraints, and $\Phi_D: \Omega \to \mathbb{R}$
is the signed distance function often used to represent implicitly the active contour
by its zero level. An early extension to image registration was proposed by 
\citep{vemuri_joint_2003} but replacing $\Phi_D$ by the intensity function of the
image to register, $\Phi_I: \Omega \to \mathbb{R}$ (the moving image). 
\citep{bertalmio_morphing_2000,vemuri_joint_2003} then define the speed function as
$\nu(\Phi_I(x,t)) = \Phi_I(x,t) - \Phi_T(x,t)$, where $\Phi_T(x)$ is the intensity
function of the target image. Finally, \citep{gorthi_active_2011} propose a different
multiphase label function as level set function ($\Phi_L$) that improves registration
results.

Generalizing the objective function $\Phi_G$, it is possible to introduce the 
dense deformation field of vectors $u: \mathbb{R}^n \to \mathbb{R}^n$ (tipically
$n = \{ 2, 3 \}$) assuming the conservation of the morphological description, that is,
$\Phi_G(x,t) = \Phi_G( x + du, t + dt ) \implies d\Phi_G(x,t) = 0$, where $d\Phi_G$ is
the total derivative of $\Phi_G$. Using the chain rule, this can be rewritten as the 
evolution equation of a vector flow:
\begin{equation}
\frac{\partial u(x,t)}{\partial t}= - \frac{\Phi_G}{\left| \nabla \Phi_G \right|} N_{\Phi_G},
\end{equation}
where $N_{\Phi_G}$ is the normal of the level set. Finally, they introduce the general
level set evolution equation into the vector flow to perform the registration task:
\begin{equation}
\frac{\partial u(x,t)}{\partial t} = - \nu( \Phi_G(x,t) ) N_{\Phi_G}.
\end{equation}
Therefore, the position of the level set function $\Phi_G$ at time $t$ is given by the
deformation field $u(x,t)$ and the initial level set function $\Phi_G(x,0)$, yielding:
\begin{equation}
\frac{\partial u(x,t)}{\partial t} = - \nu( \Phi_G(x + u(x,t), 0) ) N_{\Phi_G}.
\end{equation}

Here, we propose an active deformation field framework with a slightly different
formulation. The main differences with respect the approach proposed in
\citep{gorthi_active_2011} are: 1) there is no need for an explicit level set function
$\Phi_G$, replacing the level set gradient computation $N_{\Phi_G}$ with shape
gradients \citep{jehan-besson_dream2s:_2003,herbulot_segmentation_2006}; 2) 
\todo[inline]{regularization is also based on linear diffusion smoothing REF-thirion,
but they use a Gaussian filtering)}; 3) optimization is applied in the spectral
domain, observing anisotropic and inhomogeneous mappings along each direction.

>>>>

\Gls{acwe} and level-set based formulations have been widely applied in image
processing, as reviewed in \citep{suri_shape_2002}. We suggest clustering the 
current methodologies of template-based segmentation methods into three groups. 
The first group typically adds a shape prior term to the energy functional of 
an evolving active contour \citep{bresson_variational_2006,
chan_level_2005,chen_using_2002,cremers_kernel_2006,gastaud_combining_2004}.
These methods generally have an explicit description of the expected relative boundary 
locations of the object to be delineated, and some even model the statistical deviations
from this average shape. By including a coordinates mapping, it is possible to perform
active contours based registration between timesteps in a time-series or between different
images \citep{bertalmio_morphing_2000,wyatt_map_2003,paragios_level_2003,vemuri_joint_2003,
yezzi_variational_2003}.
A summary of these second set of techniques is performed in \citep{droske_mumfordshah_2009},
proposing two different approaches to applying the Mumford-Shah \citep{mumford_optimal_1989}
functional in joint registration and segmentation. Finally, a derivation of the 
latter groups is composed by atlas-based segmentation methods 
\citep{gorthi_segmentation_2009,gorthi_active_2011,pohl_unifying_2005,
pohl_bayesian_2006,wang_joint_2006}, where the prior 
imposes consistent voxel-based classification of contiguous regions.
A comprehensive summary of the \gls{acwe}-derived methodologies with special attention
to joint registration-segmentation methods is found in \citep{gorthi_active_2011}.

Our joint registration and segmentation method is mainly inspired in the active
deformation fields presented by \citep{gorthi_active_2011}, that is a generalization
of the methodologies under second and third groups before. As it is described in
their work, the first joint ``morphing''-segmentation approach (it was not a proper
registration method) was proposed in \citep{bertalmio_morphing_2000}. The first
registration framework is presented by \citep{yezzi_variational_2001} where the
energy functional is defined simultaneously in the moving and target images with
an affine transformation supporting the coordinates mapping (``2 PDEs approach'').
\citep{vemuri_joint_2003} proposed the first atlas-based registration based on the 
level set framework and using only one PDE. \citep{unal_coupled_2005} extended the 
idea of \citep{bertalmio_morphing_2000,yezzi_variational_2001} for non-linear 
registration with a dense deformation field as mapping function. \citep{droske_mumfordshah_2009}
combining ideas from both branches and proposed the propagation of the 
deformation field from the contours to the whole image definition. Finally,
\citep{gorthi_active_2011} present a comprehensive generalization of the methodologies.

All the surveyed frameworks are deduced from the general level set evolution equation
introduced by \citep{osher_fronts_1988}:
\begin{equation}
\frac{\partial \Phi_D(x,t)}{\partial t} = \nu ( \Phi_D(x,t)) \, \left| \nabla \Phi_D(x,t) \right|
\end{equation}
where $\nu$ is the velocity of the flow or speed function that contains the local
segmentation and contour regularization constraints, and $\Phi_D: \Omega \to \mathbb{R}$
is the signed distance function often used to represent implicitly the active contour
by its zero level. An early extension to image registration was proposed by 
\citep{vemuri_joint_2003} but replacing $\Phi_D$ by the intensity function of the
image to register, $\Phi_I: \Omega \to \mathbb{R}$ (the moving image). 
\citep{bertalmio_morphing_2000,vemuri_joint_2003} then define the speed function as
$\nu(\Phi_I(x,t)) = \Phi_I(x,t) - \Phi_T(x,t)$, where $\Phi_T(x)$ is the intensity
function of the target image. Finally, \citep{gorthi_active_2011} propose a different
multiphase label function as level set function ($\Phi_L$) that improves registration
results.

Naming $\Phi_G$ a generic objective function, a dense deformation field of vectors 
$u: \mathbb{R}^n \to \mathbb{R}^n$ (tipically $n = \{ 2, 3 \}$) is introduced. 
Then, the conservation of the morphological description is assumed, such as
$\Phi_G(x,t) = \Phi_G( x + du, t + dt ) \implies d\Phi_G(x,t) = 0$, where $d\Phi_G$ is
the total derivative of $\Phi_G$. Using the chain rule, it can be rewritten as the 
evolution equation of a vector flow:
\begin{equation}
\frac{\partial u(x,t)}{\partial t}= - \frac{\Phi_G}{\left| \nabla \Phi_G \right|} N_{\Phi_G},
\end{equation}
where $N_{\Phi_G}$ is the normal of the level set. Finally, they introduce the general
level set evolution equation into the vector flow to perform the registration task:
\begin{equation}
\frac{\partial u(x,t)}{\partial t} = - \nu( \Phi_G(x,t) ) N_{\Phi_G}.
\end{equation}
Therefore, the position of the level set function $\Phi_G$ at time $t$ is given by the
deformation field $u(x,t)$ and the initial level set function $\Phi_G(x,0)$, yielding:
\begin{equation}
\frac{\partial u(x,t)}{\partial t} = - \nu( \Phi_G(x + u(x,t), 0) ) N_{\Phi_G}.
\end{equation}

Here, we propose an active deformation field framework with some differences in
the formulation of the problem. The main diverging points with respect to
\citep{gorthi_active_2011} are: 1) there is no need for an explicit level set function
$\Phi_G$, as we replace the level set gradient computation $N_{\Phi_G}$ with shape
gradients \citep{jehan-besson_dream2s:_2003,herbulot_segmentation_2006}; 2) 
\todo[inline]{regularization is also based on linear diffusion smoothing REF-thirion,
but they use a Gaussian filtering)}; 3) optimization is applied in the spectral
domain, observing anisotropic and inhomogeneous mappings along each direction.
